---
title: "Exploring Income and Mortgage Patterns in New York"
author: "Broker T's (Team 9)"
editor: visual
format:
  html:
    code-fold: true
    toc: true         
    toc-depth: 3 
execute: 
  cache: true
  cache-rebuild: false
  warning: false
  message: false
---
```{r}
if (!require(tidyverse)) install.packages("tidyverse")
if (!require(DT)) install.packages("DT")
if (!require(sf)) install.packages("sf")
if (!require(sf)) install.packages("rvest")
if (!require(sf)) install.packages("httr")
if (!require(sf)) install.packages("data.table")

library(tidyverse)
library(DT)
library(sf)
library(utils)
library(rvest)
library(httr)
library(data.table)
```


::: {.callout .details}
<details>
<summary><h3>This section will be removed later</h3></summary>
### Overarching Question
How does the county you live in and the income you make affect mortgage rates, particularly throughout the three stages of the COVID-19 pandemic (2018-2023)?

### Specific Questions
1. Is there a county with more of a specific loan type than others? What could explain this?
2. What loan types were most popular pre, during, and post-pandemic?
3. Based on race and sex, who took out the most and least loans? Were specific loans dominant?
4. Is there a skew in the total number of loans given out depending on the county or geographic area?
5. Do interest rates vary significantly by county, and how are they influenced by sex or race?

### Data Sources
1. **New York shapefile** 
   Sourse: [US Census Bureau](https://www2.census.gov/geo/tiger/TIGER2024/CD/)

2. **NYC Annualized Sales (2003-2023)**  
   Source: [NYC Department of Finance](https://www.nyc.gov/site/finance/property/property-annualized-sales-update.page#)  
   Description: Provides property sales data for NYC.

3. **Record Mortgages Originated NY (2016-2021)**  
   Source: [NY State Department of Financial Services](https://www.dfs.ny.gov/apps_and_licensing/mortgage_companies/rrpm_originated_nys)  
   Description: Contains records of mortgages originated in New York State.

4. **HMDA US Mortgage Data (5 Years)**  
   Source: [Consumer Financial Protection Bureau](https://www.consumerfinance.gov/data-research/hmda/)  
   Description: Loan-level mortgage data, including interest rates, loan terms, and loan types.

5. **S1901: Income in the Past 12 Months**  
   Source: [US Census Bureau](https://data.census.gov/profile/New_York?g=040XX00US36)  
   Description: Annual income data by household type and income brackets.

6. **S2401: Occupation by Sex for the Civilian Employed Population 16 Years and Over**  
   Source: [US Census Bureau](https://data.census.gov/profile/New_York?g=040XX00US36)  
   Description: Employment statistics by occupation and demographics.

 
   
### Project Goals
- Examine the relationship between income, geographic location, and mortgage trends.
- Analyze demographic factors (race, sex) influencing mortgage patterns.
- Study the impact of the COVID-19 pandemic on mortgage rates and types.

### Team Members
- **Maria Jerez**
- **Lauren Cardieri**
- **Ayrat Aymetov**

### Folder Structure
- `/data`: Contains raw and cleaned datasets.
- `/scripts`: Includes code for data analysis and visualization.
- `/docs`: Documentation related to the project.
</details>
:::

## Data downloading

### New York shapefile

#### Downloading and opening New York shapefile
```{r}
# Define the URL of the ZIP file and the path for the local file
zip_url <- "https://www.example.com/tl_2024_36_cd119.zip" # Replace with actual URL
zip_file <- "data/tl_2024_36_cd119.zip"
extract_dir <- "data/tl_2024_36_cd119"

# Check if the ZIP file already exists
if (!file.exists(zip_file)) {
  message("Downloading ZIP file...")
  download.file(zip_url, zip_file, mode = "wb")
}

# Check if the extracted folder exists; if not, unzip the file
if (!dir.exists(extract_dir)) {
  message("Unzipping the file...")
  unzip(zipfile = zip_file, exdir = extract_dir)
} 

# Load the shapefile (adjust path as needed)
shapefile_path <- file.path(extract_dir, "tl_2024_36_cd119.shp")
shapefile <- st_read(shapefile_path, quiet = TRUE)

# Plot the map
plot(st_geometry(shapefile))
```

```{r}
# Define file paths
zip_file <- "data/nyc_borough_boundaries.zip"
unzip_dir <- "data/nyc_borough_boundaries"

# Download the NYC borough boundaries shapefile if not already downloaded
if (!file.exists(zip_file)) {
  download.file("https://data.cityofnewyork.us/api/geospatial/tqmj-j8zm?method=export&format=Shapefile", 
                destfile = zip_file, mode = "wb")
}

# Unzip the file
unzip(zip_file, exdir = unzip_dir)

# Find the .shp file in the unzipped directory
shp_file <- list.files(unzip_dir, pattern = "\\.shp$", full.names = TRUE)

# Load the shapefile using sf
nyc_sf <- st_read(shp_file, quiet = TRUE)

# Plot the shapefile using ggplot2
ggplot(nyc_sf) + 
  geom_sf(aes(fill = shape_area)) + 
  labs(title = "NYC Borough Boundaries", fill = "Area") +
  theme_minimal()
```


### NYC Annualized Sales (2003-2023)
```{r}
# Create the target directory if it does not exist
dir.create("data/NYC_Annualized_Sales", recursive = TRUE, showWarnings = FALSE)

# Define the base URL
base_url <- "https://www.nyc.gov/assets/finance/downloads/pdf/rolling_sales/neighborhood_sales/"

# Define the list of file names and relative paths (same strings used for both)
files <- c(
  "2023/2023_manhattan.xlsx",
  "2023/2023_bronx.xlsx",
  "2023/2023_brooklyn.xlsx",
  "2023/2023_queens.xlsx",
  "2023/2023_statenisland.xlsx",
  "2023/2023_citywide_sale.xlsx",
  "2022/2022_citywide_sale.xlsx",
  "2021_citywide_sale.xlsx",
  "2020_citywide_sale.xlsx",
  "2019_citywide_sale.xlsx",
  "2018_citywide_sale.xlsx"
)

# Use purrr::walk for mapping
walk(files, ~ {
  # Destination file name (extracts the last part of the file path)
  dest_file <- file.path("data/NYC_Annualized_Sales", basename(.x))
  
  # Check if file exists
  if (!file.exists(dest_file)) {
    tryCatch({
      # Concatenate base URL with the relative path
      download.file(paste0(base_url, .x), destfile = dest_file, mode = "wb")
      message("Downloaded: ", dest_file)
    }, error = function(e) {
      message("Failed to download: ", dest_file, " - ", e$message)
    })
  } else {
    message("File already exists: ", dest_file)
  }
})

```

### Record Mortgages Originated NY (2016-2021)
```{r}
# URL of the webpage
url <- "https://www.dfs.ny.gov/apps_and_licensing/mortgage_companies/rrpm_originated_nys"

# Define the output file path
output_file <- "data/mortgage_table.csv"

# Check if the file already exists
if (file.exists(output_file)) {
  # Load the existing CSV file into table_df
  table_df <- read.csv(output_file)
} else {
  # Read the webpage content
  webpage <- read_html(url)
  
  # Extract the table (assumes the table is the first one on the page)
  table <- webpage %>%
    html_node("table") %>%  # Select the table element
    html_table()
  
  # Convert to a data frame
  table_df <- as.data.frame(table)
  
  # Save the table as a CSV file
  write.csv(table_df, output_file, row.names = FALSE)
}

# Use table_df for further processing or display
table_df

```

### HMDA US Mortgage Data (2018-2023)

```{r}
# Base URL
base_url <- "https://ffiec.cfpb.gov/v2/data-browser-api/view/csv"

# Define years
years <- 2018:2023

# Create the output folder
dir.create("data/HMDA_NYC_Mortgage_2018_2023", recursive = TRUE, showWarnings = FALSE)

# Loop through years to download state-level data
for (year in years) {
  # Construct the API URL with states only
  api_url <- paste0(base_url, "?years=", year, "&states=36")
  
  # Output file path
  output_file <- paste0("data/HMDA_NYC_Mortgage_2018_2023/mortgage_data_", year, "_state.csv")
  
  # Check if the file already exists
  if (!file.exists(output_file)) {
    # If the file doesn't exist, download it
    tryCatch({
      response <- GET(api_url)
      
      # Check if the request is successful
      if (status_code(response) == 200) {
        writeBin(content(response, "raw"), output_file)
        message("Downloaded: ", output_file)
      } else {
        message("Failed for year ", year, " - Status Code: ", status_code(response))
        message(content(response, as = "text"))
      }
    }, error = function(e) {
      message("Error downloading year ", year, " - ", e$message)
    })
  }
}

```
```{r}
# Define the file path
file_path <- "data/HMDA_NYC_Mortgage_2018_2023/mortgage_data_2018_state.csv"

# Read the CSV file
mortgage_data <- fread(file_path)

# Get a summary of the data
colnames(mortgage_data)
```

### S1901: Income in the Past 12 Months

### S2401: Occupation by Sex for the Civilian Employed Population 16 Years and Over**  
